---
title: "Visium Example"
author: "Dylan Cable and Gabriel Grajeda"
date: "Feb 12th, 2025"
output:  
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{visium-example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    out.width = "100%"
)
```

## Installation

You can install the latest version of spacexr from Bioconductor:

```{r installation, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("spacexr")
```

## Introduction

Robust Cell Type Decomposition (RCTD) is a statistical method for decomposing
cell type mixtures in spatial transcriptomics data. In this vignette, we will
use an example Visium dataset to demonstrate how you can run RCTD on spatial
transcriptomics data and visualize your results.

We will use the term *pixel*, synonymous with *spot* or *bead*, to refer to any
observation that measures gene expression at a fixed location. We do not use
the term *cell* because, in spatial transcriptomics data, single *pixels* may
contain mixtures of multiple cells. The purpose of RCTD is to estimate the
proportions of different cell types in each pixel of a dataset.

## Setup

```{r setup, message = FALSE}
library(ggplot2)
library(SummarizedExperiment)

library(spacexr)
```

## Data Preprocessing

Before running RCTD, we must first load in:

- an annotated reference dataset, which is used to learn cell type profiles.
- a spatial transcriptomics dataset, which contains pixels for deconvolution.

### Reference

The reference is created using the `Reference` constructor function, which takes
the following parameters:

*Required parameters*:

1. `counts`: A matrix (or `dgCMatrix`) of gene expression counts, with genes as
rows and cells as columns. Row names should be unique gene names, and column
names should be unique cell barcodes. Counts should be untransformed integers.

2. `cell_types`: A factor vector containing cell type annotations for each cell
in the reference (identified by barcode). The factor levels should represent the
possible cell types.

*Optional parameters*:

3. `nUMI`: A numeric vector of total UMI counts per cell (identified by
barcode). If not provided, `nUMI` will be calculated as the column sums of the
counts matrix.

4. `min_UMI`: Minimum UMI count for cells to be included in the reference. Cells
with fewer UMIs will be removed.

5. `n_max_cells`: Maximum number of cells per cell type. If exceeded, cells will
be randomly downsampled to this number.

In practice, your reference could be an annotated dataset from:

- Single-nucleus RNA sequencing (snRNA-seq)
- Single-cell RNA sequencing (scRNA-seq)
- Cell type-specific bulk RNA sequencing

```{r reference}
refdir <- system.file("extdata",'Reference/Visium_Ref',package = 'spacexr') # directory for the reference
counts <- read.csv(file.path(refdir,"counts.csv")) # load in cell types
rownames(counts) <- counts[,1]; counts[,1] <- NULL # Move first column to rownames
cell_types <- read.csv(file.path(refdir,"cell_types.csv")) # load in cell types
cell_types <- setNames(cell_types[[2]], cell_types[[1]])
cell_types <- as.factor(cell_types) # convert to factor data type
nUMI <- read.csv(file.path(refdir,"nUMI.csv")) # load in cell types
nUMI <- setNames(nUMI[[2]], nUMI[[1]])
reference <- Reference(counts, cell_types, nUMI)

# Examine reference. (optional)
print(dim(spacexr::counts(reference))) # Gene expression matrix dimensions
table(cell_types(reference)) # Number of occurrences for each cell type
```

### Spatial Transcriptomics Data

Next, we will load the spatial transcriptomics data into a `SpatialRNA` object.
The `SpatialRNA` constructor function takes the following parameters:

*Required parameters*:

1. `coords`: A data frame (or matrix) containing x and y coordinates for each
pixel. Row names must be unique pixel barcodes, and there must be two columns
named `"x"` and `"y"` containing numeric coordinates.

2. `counts`: A matrix (or `dgCMatrix`) of gene expression counts, with genes as
rows and pixels as columns. Row names should be unique gene names, and column
names should be unique pixel barcodes. Counts should be untransformed integers.

*Optional parameters*:

3. `nUMI`: A numeric vector of total UMI counts per pixel (identified by
barcode). If not provided, `nUMI` will be calculated as the column sums of the
counts matrix.

4. `use_fake_coords`: Logical, FALSE by default. If true, the `coords` parameter
will be ignored and replaced with a placeholder coords matrix (useful for
testing).

```{r spatial, fig.align='center', fig.width=5, fig.height=5, out.width='50%'}
datadir <- system.file("extdata",'SpatialRNA/VisiumVignette',package = 'spacexr') # directory for sample Visium dataset

counts <- as.data.frame(readr::read_csv(file.path(datadir,"counts.csv"))) # load in counts matrix
coords <- read.csv(file.path(datadir,"coords.csv"))
rownames(counts) <- counts[,1]; counts[,1] <- NULL # Move first column to rownames
rownames(coords) <- coords[,1]; coords[,1] <- NULL # Move first column to rownames
nUMI <- colSums(counts) # In this case, total counts per pixel is nUMI
puck <- SpatialRNA(coords, counts, nUMI)

# Examine puck. (optional)
print(dim(spacexr::counts(puck))) # Gene expression matrix dimensions
ggplot(coords(puck), aes(x, y)) + geom_point(size = 5) + 
    theme(axis.text = element_blank(), axis.ticks = element_blank())
```

## Running RCTD

### Creating RCTD Object

The last step before running RCTD is to create an `RCTD` object from our
`SpatialRNA` and `Reference` objects using the `create.RCTD` function. Some
important configuration options include:

- `max_cores`: Number of cores used for parallel processing. We recommend
setting this to at least 4 to improve efficiency. If set to 1, parallel
processing is not used.
- `gene_cutoff`, `fc_cutoff`: Used for filtering genes in platform effect
normalization. `gene_cutoff` filters for average expression and `fc_cutoff`
filters for log-fold-change across cell types.
- `gene_cutoff_reg`, `fc_cutoff_reg`: Similar to above, but for the RCTD step.
- `UMI_min`, `UMI_max`: Minimum and maximum read depth for pixels in the
`SpatialRNA` dataset.
- `MAX_MULTI_TYPES`: Maximum number of cell types per pixel (multi mode only).

```{r createRCTD, message = FALSE}
rctd_config <- create.RCTD(puck, reference, max_cores = 4)
```

### Running RCTD

We are now ready to run RCTD using the `run.RCTD` function. 

RCTD has three modes, each suited for different spatial technologies:

- `"doublet"`: Assigns 1-2 cell types per pixel. Best for high-resolution
technologies like Slide-seq and MERFISH.
- `"full"`: No restrictions on number of cell types. Best for lower-resolution
technologies like 100-micron Visium.
- `"multi"`: Like doublet mode but can assign more cell types (up to
`MAX_MULTI_TYPES`).

The `doublet_mode` argument is used to specify the RCTD mode. For this vignette,
we'll decompose the data into doublets.

```{r runRCTD, message = FALSE}
results_se <- run.RCTD(rctd_config, doublet_mode = "doublet")
```

## RCTD Results

RCTD returns a `SummarizedExperiment` object containing:

- Three assays (one in full mode):
    1. `weights`: Cell type proportions restricted according to the specified
    mode (not available in full mode)
    2. `weights_unconfident`: Cell type proportions restricted according to the
    specified mode, including unconfident predictions (not available in full
    mode)
    2. `weights_full`: Unrestricted cell type proportions

- Additional rowData including:
    * `x`, `y`: Spatial coordinates for each pixel
    * For doublet mode:
        * `spot_class`: Classification as `singlet`, `doublet_certain`,
        `doublet_uncertain`, or `reject`
        * `first_type`, `second_type`: Predicted cell types
        * Additional metrics like `min_score`, `singlet_score` for advanced
        users
    * For multi mode:
        * `cell_type_list`: List of cell types per pixel
        * `conf_list`: List of whether cell type predictions are confident
        * Additional metrics like `min_score` for advanced users

We can view the doublet weights for our data. Each row represents a pixel, and
each column represents a cell type. The values indicate the estimated proportion
(from 0 to 1) of each cell type in that pixel:

```{r doubletWeights}
head(assay(results_se, "weights"))
```

You can access the full, unrestricted weights using
`assay(results_se, "weights_full")`. We will demonstrate this later. You may use
`assay(results_se, "weights_unconfident")` to access the less confident
predictions (e.g., the `second_type` weight for singlets and uncertain
doublets).

Below, we can view the results of RCTD deconvolution:

```{r classification}
classification_df <- data.frame(
    pixel = rownames(assay(results_se)),
    spot_class = rowData(results_se)$spot_class,
    first_type = rowData(results_se)$first_type,
    second_type = rowData(results_se)$second_type
)
head(classification_df)
```

In particular, this data frame provides us with the following information:

- `spot_class`: A factor variable with the following levels of classification:
    * `singlet` (1 cell type in the pixel)
    * `doublet_certain` (2 cell types in the pixel)
    * `doublet_uncertain` (2 cell types in the pixel, but only confident of 1)
    * `reject` (no prediction given for pixel)

    Typically, `reject` pixels should not be used, and `doublet_uncertain`
    pixels should only be used for applications that do not require knowledge of
    all cell types in a pixel.
- `first_type`: the first cell type predicted on the bead (for all spot_class
conditions except `reject`).
- `second_type`: the second cell type predicted on the bead for doublet
`spot_class` conditions (not a confident prediction for `doublet_uncertain`).

We can visualize the results of our RCTD run with the `plot_all_weights`
function. Simply pass in the RCTD results (and some aesthetic arguments), and
the cell type proportions will be plotted as pie charts for each pixel.

```{r doubletVisualization, fig.height = 5, fig.width = 7}
plot_all_weights(results_se, r = 50, lwd = 0, title = "Doublet Results")
```

We can visualize the full, unrestricted proportions by passing in the name of
the relevant assay (`"weights_full"`) to the `plot_all_weights` function:

```{r fullVisualization, fig.height = 5, fig.width = 7}
plot_all_weights(
    results_se, "weights_full", r = 50, lwd = 0, title = "Full Results"
)
```

Finally, we can visualize how the proportion of a specific cell type varies
across space with the `plot_cell_type_weight` function. We show an example with
denate.

```{r densityVisualization, fig.height = 5, fig.width = 7}
plot_cell_type_weight(
    results_se, "Denate", "weights_full",
    size = 5, stroke = 0.5, title = "Density of Denate"
)
```

You are now ready to apply RCTD to your own spatial transcriptomics datasets!

## Session Information
```{r sessionInfo}
sessionInfo()
```
